---
title: "Intra- and interspecific variation in Swiss ants"
subtitle: "Trait-trait and trait-environment correlations"
author: "Tim Szewczyk"
output:
  html_document:
    theme: spacelab
    df_print: paged
    anchor_sections: TRUE
    toc: yes
    toc_depth: 2
    toc_float: true
  pdf_document:
    toc: yes
  html_notebook:
    theme: spacelab
    toc: yes
editor_options: 
  chunk_output_type: console
runtime: shiny
---


# Overview  

```{r setup_workspace, include=FALSE, message=FALSE, warning=FALSE}
# libraries, functions, directories
library(tidyverse); library(sf); library(shiny); library(GGally)
theme_set(theme_bw() + theme(panel.grid.minor=element_blank(),
                             panel.grid.major=element_line(colour="gray90",
                                                           size=0.25)))
#' Plot points and linear smoother in GGally
ggally_pts_smooth <- function(data, mapping, method="lm", se=F,
                              size_pt=0.4, size_smooth=0.3,
                              alpha_pt=0.4, alpha_smooth=1, ...) {
  p <- ggplot(data=data, mapping=mapping) + 
    geom_point(shape=1, size=size_pt, alpha=alpha_pt) + 
    geom_smooth(method=method, formula=y~x, se=se, 
                size=size_smooth, alpha=alpha_smooth)
  p
}

min_clny_size <- 3

trait_match <- data.frame(orig=c("v", 
                                 "HeadLength", "HeadWidth", 
                                 "InterocularDistance", "DVE",
                                 "ScapeLength", "WebersLength",
                                 "HindTibia", "MidTibia", "MesosomaWidth",
                                 "MesosomaLength", "HindFemur", "MidFemur",
                                 "MidLen", "HindLen", "MesoSA", 
                                 "HeadShape", "PronotExp", "ScapeProp"),
                          full=c("Color (lightness)", 
                                 "Head Length", "Head Width",
                                 "Interocular Dist. (rel)", 
                                 "Dorsoventral Eye Pos.",
                                 "Scape Length", "Webers Length",
                                 "Hind Tibia", "Mid Tibia", "Pronotum Width",
                                 "Mesosoma Length", "Hind Femur", "Mid Femur",
                                 "Rel. Leg Length (mid)", 
                                 "Rel. Leg Length (hind)", 
                                 "Mesosoma SA", "Head Shape",
                                 "Pronotal Expansion", "Scape Proportion"),
                          relative=c(F, 
                                     F, F,
                                     T, T, 
                                     F, F, 
                                     F, F, F, 
                                     F, F, F,
                                     T, T, F, 
                                     T, T, T))
focus_traits <- c("Scape Proportion",
                  "Dorsoventral Eye Pos.", 
                  "Rel. Leg Length (hind)",
                  "Pronotal Expansion",
                  "Head Shape",
                  "Head Width", 
                  "Head Length",
                  "Webers Length", 
                  "Color (lightness)")

trts <- readRDS("trts.rds")
trts$clny.wide <- filter(trts$clny.wide, npp > 0)

wkr.df <- filter(trts$wkr.wide, n_clny >= min_clny_size) %>% 
  select(-`NA`) %>%
  mutate(MidLen=MidLen/WebersLength, HindLen=HindLen/WebersLength,
         DVE=InterocularDistance/HeadWidth,
         InterocularDistance=(HeadWidth-InterocularDistance)/HeadLength,
         HeadShape=HeadWidth/HeadLength,
         PronotExp=MesosomaWidth/WebersLength,
         ScapeProp=ScapeLength/HeadLength) %>%
  select(one_of("Worker", "TubeNo", "SPECIESID", trait_match$orig)) %>%
  pivot_longer(cols=3+(seq_along(trait_match$full)), 
               names_to="Trait", values_to="Value") %>%
  group_by(Trait, SPECIESID) %>% 
  mutate(Value_std=scale(Value)[,1]) %>%
  ungroup %>%
  mutate(GENUSID=str_sub(SPECIESID, 1, 4),
         Trait_orig=Trait,
         Trait=trait_match$full[match(Trait, trait_match$orig)]) %>%
  filter(!is.na(Value))
```

```{r complete_df_setup, echo=F, message=F, warning=F}
wkr.df_wide <- wkr.df %>% 
  select(-Value_std, -Trait_orig) %>%
  filter(Trait %in% focus_traits) %>%
  pivot_wider(names_from="Trait", values_from=c("Value")) %>%
  mutate(el=trts$clny.wide$mnt25[match(TubeNo, trts$clny.wide$TubeNo)])
complete.df <- wkr.df_wide[complete.cases(wkr.df_wide),]
complete.clny <- complete.df %>% group_by(GENUSID, SPECIESID, TubeNo) %>%
  summarise(across(where(is.numeric), 
                   list(mn=mean, sd=~log(sd(.x)), CV=~log(sd(.x))/mean(.x)))) %>%
  ungroup
```



# Correlations

## Trait-trait {.tabset .tabset-fade}  

Be patient with these. The plots are made with the package `GGally` and it isn't the fastest. The first time a tab is opened, `shiny` will display an error until a graph can be produced with the default inputs.

### Worker: y

```{r wkr_trt_trt, echo=FALSE, message=F, warning=F, fig.height=10}
shinyApp(
  ui=fluidPage(
    fluidRow(
      column(width=12,
             selectInput(inputId="wkrTT",
                         label="Traits",
                         choices=as.list(focus_traits),
                         selected=focus_traits[c(3,8,9)], 
                         multiple=T))),
    fluidRow(
      plotOutput("wkrTT.ggally")
    )
  ),
  server=function(input, output) {
    wkrTT.df <- reactive({
      complete.df %>% select(SPECIESID, GENUSID, any_of(input$wkrTT))
    })
    output$wkrTT.ggally <- renderPlot({
      wkrTT.df() %>% 
        ggpairs(columns=3:ncol(.), 
                aes(colour=GENUSID, group=SPECIESID, fill=GENUSID),
                lower=list(continuous=wrap(ggally_pts_smooth)),
                diag=list(continuous=wrap('densityDiag', alpha=0.1, size=0.25)),
                upper=list(continuous=wrap(ggally_cor, digits=2, size=3, 
                                           family='serif', title="Total")),
                labeller=label_wrap_gen(15)) + 
        scale_colour_brewer(type="qual", palette=2) +
        scale_fill_brewer(type="qual", palette=2) +
        theme(panel.grid.minor=element_blank(), 
              panel.grid.major=element_line(colour="grey95", size=0.1), 
              axis.text=element_blank(), axis.ticks=element_blank(),
              legend.position="none")
    }, height=450, width=450)
  },
  options=list(height=555)
)
```

### Colony mean: $\bar{y}$

```{r ybar_trt_trt, echo=FALSE, message=F, warning=F, fig.height=9}
shinyApp(
  ui=fluidPage(
    fluidRow(
      column(width=12,
             selectInput(inputId="ybarTT",
                         label="Traits",
                         choices=as.list(focus_traits),
                         selected=focus_traits[c(3,8,9)],
                         multiple=T))),
    fluidRow(
      plotOutput("ybarTT.ggally")
    )
  ),
  server=function(input, output) {
    ybarTT.df <- reactive({
      complete.clny %>% 
        select(SPECIESID, GENUSID, any_of(paste0(input$ybarTT, "_mn"))) %>%
        rename_with(~str_remove(.x, "_mn"), contains("_mn"))
    })
    output$ybarTT.ggally <- renderPlot({
      ybarTT.df() %>% 
        ggpairs(columns=3:ncol(.), 
                aes(colour=GENUSID, group=SPECIESID, fill=GENUSID),
                lower=list(continuous=wrap(ggally_pts_smooth)),
                diag=list(continuous=wrap('densityDiag', alpha=0.1, size=0.25)),
                upper=list(continuous=wrap(ggally_cor, digits=2, size=3, 
                                           family='serif', title="Total")),
                labeller=label_wrap_gen(15)) + 
        scale_colour_brewer(type="qual", palette=2) +
        scale_fill_brewer(type="qual", palette=2) +
        theme(panel.grid.minor=element_blank(), 
              panel.grid.major=element_line(colour="grey95", size=0.1), 
              axis.text=element_blank(), axis.ticks=element_blank(),
              legend.position="none")
    }, height=450, width=450)
  },
  options=list(height=555)
)
```


### Colony sd: $log(d)$

```{r dlog_trt_trt, echo=FALSE, message=F, warning=F, fig.height=9}
shinyApp(
  ui=fluidPage(
    fluidRow(
      column(width=12,
             selectInput(inputId="dlogTT",
                         label="Traits",
                         choices=as.list(focus_traits),
                         selected=focus_traits[c(3,8,9)],
                         multiple=T))),
    fluidRow(
      plotOutput("dlogTT.ggally")
    )
  ),
  server=function(input, output) {
    dlogTT.df <- reactive({
      complete.clny %>% 
        select(SPECIESID, GENUSID, any_of(paste0(input$dlogTT, "_sd"))) %>%
        rename_with(~str_remove(.x, "_sd"), contains("_sd"))
    })
    output$dlogTT.ggally <- renderPlot({
      dlogTT.df() %>% 
        ggpairs(columns=3:ncol(.), 
                aes(colour=GENUSID, group=SPECIESID, fill=GENUSID),
                lower=list(continuous=wrap(ggally_pts_smooth)),
                diag=list(continuous=wrap('densityDiag', alpha=0.1, size=0.25)),
                upper=list(continuous=wrap(ggally_cor, digits=2, size=3, 
                                           family='serif', title="Total")),
                labeller=label_wrap_gen(15)) + 
        scale_colour_brewer(type="qual", palette=2) +
        scale_fill_brewer(type="qual", palette=2) +
        theme(panel.grid.minor=element_blank(), 
              panel.grid.major=element_line(colour="grey95", size=0.1), 
              axis.text=element_blank(), axis.ticks=element_blank(),
              legend.position="none")
    }, height=450, width=450)
  },
  options=list(height=555)
)
```


### Colony CV: $log(d)/\bar{y}$

```{r CV_trt_trt, echo=FALSE, message=F, warning=F, fig.height=9}
shinyApp(
  ui=fluidPage(
    fluidRow(
      column(width=12,
             selectInput(inputId="CVTT",
                         label="Traits",
                         choices=as.list(focus_traits),
                         selected=focus_traits[c(3,8,9)],
                         multiple=T))),
    fluidRow(
      plotOutput("CVTT.ggally")
    )
  ),
  server=function(input, output) {
    CVTT.df <- reactive({
      complete.clny %>% 
        select(SPECIESID, GENUSID, any_of(paste0(input$CVTT, "_CV"))) %>%
        rename_with(~str_remove(.x, "_CV"), contains("_CV"))
    })
    output$CVTT.ggally <- renderPlot({
      CVTT.df() %>% 
        ggpairs(columns=3:ncol(.), 
                aes(colour=GENUSID, group=SPECIESID, fill=GENUSID),
                lower=list(continuous=wrap(ggally_pts_smooth)),
                diag=list(continuous=wrap('densityDiag', alpha=0.1, size=0.25)),
                upper=list(continuous=wrap(ggally_cor, digits=2, size=3, 
                                           family='serif', title="Total")),
                labeller=label_wrap_gen(15)) + 
        scale_colour_brewer(type="qual", palette=2) +
        scale_fill_brewer(type="qual", palette=2) +
        theme(panel.grid.minor=element_blank(), 
              panel.grid.major=element_line(colour="grey95", size=0.1), 
              axis.text=element_blank(), axis.ticks=element_blank(),
              legend.position="none")
    }, height=450, width=450)
  },
  options=list(height=555)
)
```








## Environment-trait {.tabset .tabset-fade}

### Worker: y

```{r wkr_env_trt, echo=FALSE, message=F, warning=F}
shinyApp(
  ui=fluidPage(
    fluidRow(
      column(width=6,
             selectInput(inputId="wkrE",
                  label="Environmental variables",
                  choices=as.list(names(trts$clny.wide)[c(13,18:21,23:24,27:30)]),
                  selected=c("TwarmQ", "PwarmQ", "npp", "SampleDate"),
                  multiple=T)),
      column(width=3, 
             selectInput(inputId="wkrT",
                  label="Traits",
                  choices=as.list(focus_traits),
                  selected=focus_traits[1]))),
    fluidRow(
      plotOutput("wkrET.ggally")
    )
  ),
  server=function(input, output) {
    wkrET.df <- reactive({
      complete.df %>% select(TubeNo, SPECIESID, GENUSID, any_of(input$wkrT)) %>%
        left_join(., select(trts$clny.wide, TubeNo, any_of(input$wkrE)), 
                   by="TubeNo") %>%
        select(SPECIESID, GENUSID, any_of(input$wkrE), any_of(input$wkrT))
    })
    output$wkrET.ggally <- renderPlot({
      wkrET.df() %>% 
        ggpairs(columns=3:ncol(.), 
                aes(colour=GENUSID, group=SPECIESID, fill=GENUSID),
                lower=list(continuous=wrap(ggally_pts_smooth)),
                diag=list(continuous=wrap('densityDiag', alpha=0.1, size=0.25)),
                upper=list(continuous=wrap(ggally_cor, digits=2, size=3, 
                                           family='serif', title="Total")),
                labeller=label_wrap_gen(15)) + 
        scale_colour_brewer(type="qual", palette=2) +
        scale_fill_brewer(type="qual", palette=2) +
        theme(panel.grid.minor=element_blank(), 
              panel.grid.major=element_line(colour="grey95", size=0.1), 
              axis.text=element_blank(), axis.ticks=element_blank(),
              legend.position="none")
    }, height=550, width=550)
  },
  options=list(height=650)
)
```

### Colony mean: $\bar{y}$

```{r ybar_env_trt, echo=FALSE, message=F, warning=F, fig.height=9}
shinyApp(
  ui=fluidPage(
    fluidRow(
          column(width=6,
                 selectInput(inputId="ybarE",
                  label="Environmental variables",
                  choices=as.list(names(trts$clny.wide)[c(13,18:21,23:24,27:30)]),
                  selected=c("TwarmQ", "PwarmQ", "npp", "SampleDate"),
                  multiple=T)),
          column(width=3, 
             selectInput(inputId="ybarT",
                  label="Traits",
                  choices=as.list(focus_traits),
                  selected=focus_traits[1]))),
    fluidRow(
      plotOutput("ybarET.ggally")
    )
  ),
  server=function(input, output) {
    ybarET.df <- reactive({
      complete.clny %>% 
        select(TubeNo, SPECIESID, GENUSID, any_of(paste0(input$ybarT, "_mn"))) %>%
        rename_with(~str_remove(.x, "_mn"), contains("_mn")) %>%
        left_join(., select(trts$clny.wide, TubeNo, any_of(input$ybarE)), 
                   by="TubeNo") %>%
        select(SPECIESID, GENUSID, any_of(input$ybarE), any_of(input$ybarT))
    })
    output$ybarET.ggally <- renderPlot({
      ybarET.df() %>% 
        ggpairs(columns=3:ncol(.), 
                aes(colour=GENUSID, group=SPECIESID, fill=GENUSID),
                lower=list(continuous=wrap(ggally_pts_smooth)),
                diag=list(continuous=wrap('densityDiag', alpha=0.1, size=0.25)),
                upper=list(continuous=wrap(ggally_cor, digits=2, size=3, 
                                           family='serif', title="Total")),
                labeller=label_wrap_gen(15)) + 
        scale_colour_brewer(type="qual", palette=2) +
        scale_fill_brewer(type="qual", palette=2) +
        theme(panel.grid.minor=element_blank(), 
              panel.grid.major=element_line(colour="grey95", size=0.1), 
              axis.text=element_blank(), axis.ticks=element_blank(),
              legend.position="none")
    }, height=550, width=550)
  },
  options=list(height=650)
)
```


### Colony sd: $log(d)$

```{r dlog_env_trt, echo=FALSE, message=F, warning=F, fig.height=9}
shinyApp(
  ui=fluidPage(
    fluidRow(
      column(width=6,
                 selectInput(inputId="dlogE",
                  label="Environmental variables",
                  choices=as.list(names(trts$clny.wide)[c(13,18:21,23:24,27:30)]),
                  selected=c("TwarmQ", "PwarmQ", "npp", "SampleDate"),
                  multiple=T)),
      column(width=3, 
             selectInput(inputId="dlogT",
                  label="Traits",
                  choices=as.list(focus_traits),
                  selected=focus_traits[1]))),
    fluidRow(
      plotOutput("dlogET.ggally")
    )
  ),
  server=function(input, output) {
    dlogET.df <- reactive({
      complete.clny %>% 
        select(TubeNo, SPECIESID, GENUSID, any_of(paste0(input$dlogT, "_sd"))) %>%
        rename_with(~str_remove(.x, "_sd"), contains("_sd")) %>%
        left_join(., select(trts$clny.wide, TubeNo, any_of(input$dlogE)), 
                   by="TubeNo") %>%
        select(SPECIESID, GENUSID, any_of(input$dlogE), any_of(input$dlogT))
    })
    output$dlogET.ggally <- renderPlot({
      dlogET.df() %>% 
        ggpairs(columns=3:ncol(.), 
                aes(colour=GENUSID, group=SPECIESID, fill=GENUSID),
                lower=list(continuous=wrap(ggally_pts_smooth)),
                diag=list(continuous=wrap('densityDiag', alpha=0.1, size=0.25)),
                upper=list(continuous=wrap(ggally_cor, digits=2, size=3, 
                                           family='serif', title="Total")),
                labeller=label_wrap_gen(15)) + 
        scale_colour_brewer(type="qual", palette=2) +
        scale_fill_brewer(type="qual", palette=2) +
        theme(panel.grid.minor=element_blank(), 
              panel.grid.major=element_line(colour="grey95", size=0.1), 
              axis.text=element_blank(), axis.ticks=element_blank(),
              legend.position="none")
    }, height=550, width=550)
  },
  options=list(height=650)
)
```


### Colony CV: $log(d)/\bar{y}$

```{r CV_env_trt, echo=FALSE, message=F, warning=F, fig.height=9}
shinyApp(
  ui=fluidPage(
    fluidRow(
      column(width=6,
                 selectInput(inputId="CVE",
                  label="Environmental variables",
                  choices=as.list(names(trts$clny.wide)[c(13,18:21,23:24,27:30)]),
                  selected=c("TwarmQ", "PwarmQ", "npp", "SampleDate"),
                  multiple=T)),
      column(width=3, 
             selectInput(inputId="CVT",
                  label="Traits",
                  choices=as.list(focus_traits),
                  selected=focus_traits[1]))),
    fluidRow(
      plotOutput("CVET.ggally")
    )
  ),
  server=function(input, output) {
    CVET.df <- reactive({
      complete.clny %>% 
        select(TubeNo, SPECIESID, GENUSID, any_of(paste0(input$CVT, "_CV"))) %>%
        rename_with(~str_remove(.x, "_CV"), contains("_CV")) %>%
        left_join(., select(trts$clny.wide, TubeNo, any_of(input$CVE)), 
                   by="TubeNo") %>%
        select(SPECIESID, GENUSID, any_of(input$CVE), any_of(input$CVT))
    })
    output$CVET.ggally <- renderPlot({
      CVET.df() %>% 
        ggpairs(columns=3:ncol(.), 
                aes(colour=GENUSID, group=SPECIESID, fill=GENUSID),
                lower=list(continuous=wrap(ggally_pts_smooth)),
                diag=list(continuous=wrap('densityDiag', alpha=0.1, size=0.25)),
                upper=list(continuous=wrap(ggally_cor, digits=2, size=3, 
                                           family='serif', title="Total")),
                labeller=label_wrap_gen(15)) + 
        scale_colour_brewer(type="qual", palette=2) +
        scale_fill_brewer(type="qual", palette=2) +
        theme(panel.grid.minor=element_blank(), 
              panel.grid.major=element_line(colour="grey95", size=0.1), 
              axis.text=element_blank(), axis.ticks=element_blank(),
              legend.position="none")
    }, height=550, width=550)
  },
  options=list(height=650)
)
```

