---
title: "Intra- and interspecific variation in Swiss ants"
subtitle: "Distributional models in brms"
author: "Tim Szewczyk"
output:
  html_document:
    theme: spacelab
    df_print: paged
    anchor_sections: TRUE
    toc: yes
    toc_depth: 2
    toc_float: true
  pdf_document:
    toc: yes
  html_notebook:
    theme: spacelab
    toc: yes
editor_options: 
  chunk_output_type: console
---


# Overview  

A recent update to the R package `brms` includes the ability to run *distributional models*, where both the mean and standard deviation are predicted. Since this is exactly what I want to do and the author of the package is an actual statistician who works closely with the Stan development team, it's better to use the package and scrap the homemade models I had been using. This also gives the package (and Stan) some extra visibility, and gives an example of how easy-to-use Bayesian models have become.

```{r setup, echo=F}
# set directory for knitr as main project directory
knitr::opts_knit$set(root.dir=rprojroot::find_rstudio_root_file())
```

```{r setup_workspace, echo=FALSE, message=FALSE, warning=FALSE}
# libraries, functions, directories
library(tidyverse); library(sf); library(brms); library(tidybayes)
options(mc.cores=parallel::detectCores())
theme_set(theme_bw() + theme(panel.grid.minor=element_blank(),
                             panel.grid.major=element_line(colour="gray90",
                                                           size=0.25)))
googlesheets4::gs4_auth(email="tim.szewczyk@gmail.com")

source("code/00_fn.R")
walk(paste0("../1_opfo/code/", c("lc_cols", "00_fn"), ".R"), source)
trts <- readRDS("data/trts.rds")

fig_dir <- "eda/"
out_dir <- "temp/"

trait_match <- data.frame(orig=c("v", "grey_md",
                                 "HeadLength", "HeadWidth", 
                                 "RelIntOc",
                                 "ScapeLength", "WebersLength",
                                 "HindTibia", "MidTibia", "MesosomaWidth",
                                 "MesosomaLength", "HindFemur", "MidFemur",
                                 "MidLen", "HindLen", "RelLegHind", "MesoSA", 
                                 "HeadShape", "PronotExp", "ScapeProp"),
                          full=c("Color (lightness)", "Color (med. grey)",
                                 "Head Length", "Head Width",
                                 "Interocular Distance",
                                 "Scape Length", "Webers Length",
                                 "Hind Tibia", "Mid Tibia", "Pronotum Width",
                                 "Mesosoma Length", "Hind Femur", "Mid Femur",
                                 "Leg Length (mid)", 
                                 "Leg Length (hind)", "Leg Length (hind)", 
                                 "Mesosoma SA", "Head Shape",
                                 "Pronotal Expansion", "Scape Proportion"),
                          relative=c(F, F,
                                     F, F,
                                     T, 
                                     F, F, 
                                     F, F, F, 
                                     F, F, F,
                                     T, T, T, F, 
                                     T, T, T))
```



----------



# Model setup

```{r model_setup, message=FALSE, warning=FALSE}
# standardization methods:
#   1: scale cov & traits across all species
#   2: scale cov & traits within spp
#   3: scale cov across all spp, traits within spp
std_method <- 3 

# minimum number of colonies for species to be included
clny_min <- 3

# minimum elevational range for species to be included
rng_thresh <- 200

# select variable set
var_set <- c("all", "el", "climate", "hyp")[3]
fig_dir <- paste0(fig_dir, var_set, "/")

# response variables to loop through
response_vars <- c("v", 
                   # "grey_md",
                   "HeadWidth",
                   "HeadLength",
                   "HeadShape",
                   "RelIntOc",
                   # "ScapeLength",
                   "ScapeProp",
                   "WebersLength",
                   "PronotExp",
                   "RelLegHind")

# covariates: all, el, and specific hypotheses
X.ls <- list("all"=c("GDD5", "PwarmQ","Pseas", "Tseas", 
                     "npp", "SampleDate"),
             "el"=c("mnt25", "SampleDate"),
             "climate"=c("TwarmQ", "PwarmQ", "SampleDate"),
             "v"=c("TwarmQ", "PwarmQ", "SampleDate"),
             "grey_md"=c("TwarmQ", "PwarmQ", "SampleDate"),
             "HeadWidth"=c("TwarmQ", "PwarmQ", "SampleDate"),
             "HeadLength"=c("TwarmQ", "PwarmQ", "SampleDate"),
             "HeadShape"=c("TwarmQ", "PwarmQ", "SampleDate"),
             "WebersLength"=c("TwarmQ", "PwarmQ", "SampleDate"),
             "RelLegHind"=c("TwarmQ", "npp", "SampleDate"),
             "ScapeProp"=c("TwarmQ", "npp", "SampleDate"),
             "ScapeLength"=c("TwarmQ", "npp", "SampleDate"),
             "PronotExp"=c("TwarmQ", "npp", "SampleDate"),
             "RelIntOc"=c("TwarmQ", "npp", "SampleDate"))
genera_incl <- c("Myrm", 
                 "Mani",
                 "Lasi",
                 "Temn",
                 "Tetr",
                 "Lept",
                 "Form",
                 "Apha"
                 )

cov_names <- tibble(orig=c("TwarmQ", "GDD5", "PwarmQ", 
                           "Pseas", "Tseas", 
                           "SampleDate", "mnt25", "npp"),
                    full=c("Temp. (WarmQ)", "Temp. (GDD)", "Precip. (WarmQ)", 
                           "Precip. Seasonality", "Temp. Seasonality", 
                           "Date", "Elevation", "NPP"))


# wkr.df -----------------------------------------------------------------------
std_ext <- paste0("_std_", std_method)
trts$wkr.wide <- trts$wkr.wide %>%
  mutate(Cnpy_fac=factor(case_when(CnpyOpen==1 ~ "Open",
                                   CnpyMixed==1 ~ "Mixed",
                                   CnpyClosed==1 ~ "Closed")))
wkr.base <- trts$wkr.wide %>% 
  mutate(RelIntOc=InterocularDistance/HeadWidth,
         RelLegHind=HindLen/WebersLength,
         HeadShape=HeadWidth/HeadLength,
         PronotExp=MesosomaWidth/WebersLength,
         ScapeProp=ScapeLength/HeadLength)  %>%
  select(Worker, TubeNo, GENUSID, SPECIESID, SampleDate, mnt25,
         any_of(response_vars), any_of(unique(unlist(X.ls)))) 
if(std_method==1) {
  wkr.df <- wkr.base %>% 
    mutate(across(any_of(response_vars, unique(unlist(X.ls))), ~c(scale(.))))
} else if(std_method==2) {
  wkr.df <- wkr.base %>% group_by(SPECIESID) %>%
    mutate(across(any_of(response_vars, unique(unlist(X.ls))), ~c(scale(.))))
} else if(std_method==3) {
  wkr.df <- wkr.base %>% 
    mutate(across(any_of(unique(unlist(X.ls))), ~c(scale(.)))) %>%
    group_by(SPECIESID) %>%
    mutate(across(any_of(response_vars), ~c(scale(.)))) 
} 

wkr.df <- wkr.df %>% ungroup %>%
  mutate(Cnpy=trts$wkr.wide$Cnpy_fac,
         el_m=trts$wkr.wide$mnt25) %>%
  group_by(SPECIESID) %>% mutate(nColony=n_distinct(TubeNo)) %>% ungroup %>%
  filter(nColony >= clny_min) %>%
  filter(SPECIESID %in% filter(trts$spp_rng, rng >= rng_thresh)$SPECIESID) %>%
  filter(GENUSID %in% genera_incl) %>%
  arrange(SPECIESID, TubeNo) 

```



----------




# Dataset summaries

```{r data_summary, message=F, warning=F, echo=F}
trts$wkr.wide %>% summarise(nSpp=n_distinct(SPECIESID), 
                            nColonies=n_distinct(TubeNo), 
                            nWorker=n()) %>% 
  knitr::kable(caption="Total measurements")

wkr.df %>% summarise(nSpp=n_distinct(SPECIESID), 
                     nColonies=n_distinct(TubeNo), 
                     nWorker=n()) %>% 
  knitr::kable(caption="Measurements for species meeting clny_min and rng_thresh criteria")
```






----------



# Model loop

```{r model_loop, eval=F, message=F, warning=F}
for(k in 1:length(response_vars)) {
  
  Y_var <- response_vars[k]
  X_vars <- X.ls[[ifelse(var_set=="hyp", Y_var, var_set)]]
  cat("----", Y_var, "with predictors", paste(X_vars, collapse=", "), "\n")
  
  # formulas for models to compare
  form.ls <- list(
    # mn_only=bf(
    #   as.formula(paste0(Y_var, "~", 
    #                     paste(X_vars, collapse="+"),
    #                     "+ (", paste(X_vars, collapse="+"), "|SPECIESID)",
    #                     "+ (1 | TubeNo)"))
    # ),
    # mn_sd_re=bf(
    #   as.formula(paste0(Y_var, "~",
    #                     paste(X_vars, collapse="+"),
    #                     "+ (", paste(X_vars, collapse="+"), "|SPECIESID)",
    #                     "+ (1 | TubeNo)")),
    #   sigma ~ (1 | SPECIESID)
    # ),
    mn_sd_full=bf(
      as.formula(paste0(Y_var, "~", 
                        paste(X_vars, collapse="+"),
                        "+ (", paste(X_vars, collapse="+"), "|SPECIESID/GENUSID)",
                        "+ (1 | ID1 | TubeNo)")
      ),
      as.formula(paste0("sigma ~",
                        paste(X_vars, collapse="+"),
                        "+ (1 + ", paste(X_vars, collapse="+"), "|SPECIESID/GENUSID)",
                        "+ (1 | ID1 | TubeNo)"))
    ))
  sp_re <- as.formula(paste("~(", paste(X_vars, collapse="+"), "|SPECIESID)"))
  
  
  # run models
  mod.ls <- map(form.ls, ~brm(.x, 
                              prior=c(prior(normal(0, 1), class=Intercept),
                                      prior(normal(0, 1), class=b),
                                      prior(normal(0, 3), class=sd),
                                      prior(lkj(0.5), class=L)),
                              data=wkr.df, inits=0,
                              control=list(adapt_delta=0.9))) %>%
    setNames(names(form.ls))
  
  # evaluate models
  # loo_out <- loo(mod.ls[[1]], mod.ls[[2]], mod.ls[[3]],
  #                model_names=names(mod.ls))
  # as_tibble(loo_out$diffs, rownames="model") %>%
  #   write_csv(paste0(fig_dir, "loo_", Y_var, std_ext, ".csv"))
  # map_dfr(mod.ls, ~as_tibble(bayes_R2(.x)), .id="model") %>%
  #   write_csv(paste0(fig_dir, "R2_", Y_var, std_ext, ".csv"))
  # map_dfr(mod.ls, ~as_tibble(bayes_R2(.x, re_form=sp_re)), .id="model") %>%
  #   write_csv(paste0(fig_dir, "R2_spRE_", Y_var, std_ext, ".csv"))
  
  
  # plot conditional effects
  imap(mod.ls,  ~map2(plot(conditional_effects(.x, spaghetti=T, nsamples=500),
                           ask=F, plot=F, points=F, 
                           point_args=list(alpha=0.25, size=0.5, shape=1),
                           spaghetti_args=list(
                             colour=adjustcolor("cadetblue", alpha.f=0.5), 
                             size=0.05),
                           line_args=list(colour="black", size=1)), .y, 
                      ~.x + ggtitle(.y) + ylim(-1,1))) %>%
    unlist(., recursive=F) %>% 
    ggpubr::ggarrange(plotlist=., nrow=length(mod.ls), ncol=length(X_vars))
  ggsave(paste0(fig_dir, "cond_eff_", Y_var, std_ext, ".png"), 
         width=3*length(X_vars), height=3*length(mod.ls))
  
  # plot population-level estimates
  p <- map_dfr(mod.ls, ~ .x %>% 
            gather_draws(`b_.*`, regex=T) %>%
            median_hdi(.width=c(0.5, 0.8, 0.95)), 
          .id="model") %>%
    mutate(covariate=str_remove(.variable, "^(b_sigma_|b_)"),
           dpar=if_else(grepl("sigma", .variable), "sigma", "mu")) %>%
    filter(covariate != "Intercept") %>%
    ggplot(aes(y=covariate, x=.value, xmin=.lower, xmax=.upper)) +
    geom_vline(xintercept=0, colour="grey") +
    geom_pointinterval(position=position_dodge(width=0.5)) +
    # scale_colour_brewer(type="qual", palette=3, direction=-1) +
    labs(x="value", y="", title=Y_var) + 
    facet_grid(.~dpar)
  ggsave(paste0(fig_dir, "b_eff_", Y_var, std_ext, ".png"), p, 
         width=8, height=2*length(X_vars))
  
  map_dfr(mod.ls, ~ .x %>% 
            gather_draws(`b_.*`, regex=T) %>%
            median_hdi(.width=c(0.5, 0.8, 0.95)), 
          .id="model") %>%
    mutate(covariate=str_remove(.variable, "^(b_sigma_|b_)"),
           dpar=if_else(grepl("sigma", .variable), "sigma", "mu"),
           Yvar=Y_var) %>%
    write_csv(paste0(fig_dir, "b_eff_", Y_var, std_ext, ".csv"))
  
  # p <- map_dfr(mod.ls, ~ .x %>% 
  #           gather_draws(`b_.*`, regex=T), 
  #         .id="model") %>%
  #   mutate(covariate=str_remove(.variable, "^(b_sigma_|b_)"),
  #          dpar=if_else(grepl("sigma", .variable), "sigma", "mu")) %>%
  #   ggplot(aes(y=covariate, x=.value, fill=model, group=model)) +
  #   geom_vline(xintercept=0, colour="grey") +
  #   stat_slab(alpha=0.5, size=0.3, colour="grey30") +
  #   scale_colour_brewer(type="qual", palette=3, direction=-1) +
  #   scale_fill_brewer(type="qual", palette=3, direction=-1) +
  #   labs(x="value", y="", title=Y_var) + 
  #   facet_grid(.~dpar)
  # ggsave(paste0(fig_dir, "b_dens_eff_", Y_var, std_ext, ".png"), p, 
  #        width=8, height=2*length(X_vars))
  
  
  # plot species-level slopes
  b_df <- map_dfr(mod.ls, ~ .x %>% 
            gather_draws(`b_.*`, regex=T) %>%
            median_hdi(.width=c(0.5, 0.8, 0.95)), 
          .id="model") %>%
    mutate(covariate=str_remove(.variable, "^(b_sigma_|b_)"),
           dpar=if_else(grepl("sigma", .variable), "sigma", "mu"))
  r_df <- map_dfr(mod.ls, ~ .x %>% 
            gather_draws(`r_SPECIESID.*`, regex=T) %>%
            median_hdi(.width=c(0.5, 0.8, 0.95)), 
          .id="model") %>%
    mutate(covariate=str_remove(str_split_fixed(.variable, ",", 2)[,2], "]"),
           dpar=if_else(grepl("sigma", .variable), "sigma", "mu"),
           species=str_split_fixed(str_split_fixed(.variable, "\\[", 2)[,2], 
                                   ",", 2)[,1])
  spp_df <- full_join(b_df %>% select(-.variable), 
                      r_df %>% select(-.variable), 
                      by=c("model", ".point", ".width", 
                           ".interval", "covariate", "dpar"),
                      suffix=c("_b", "_r")) %>%
    mutate(.value=.value_b + .value_r,
           .lower=.lower_b + .lower_r,
           .upper=.upper_b + .upper_r) %>%
    select(-contains("_b"), -contains("_r")) %>%
    filter(!grepl("_[A-Z]", species))
  p <- ggplot(filter(spp_df, model=="mn_sd_full"), 
         aes(y=species, x=.value, xmin=.lower, xmax=.upper)) +
    geom_vline(xintercept=0, colour="grey") +
    geom_pointinterval(alpha=0.5, size=0.5, shape=1) +
    scale_fill_brewer(type="qual", palette=3, direction=-1) +
    labs(x="value", y="", title=Y_var) + 
    facet_grid(covariate~dpar)
  ggsave(paste0(fig_dir, "b_spp_eff_", Y_var, std_ext, ".png"), p, 
         width=5, height=5*length(X_vars))
  
  # plot distributional parameters
  post <- mod.ls[[1]]$data %>%
    add_fitted_draws(mod.ls[[1]], dpar=T) %>%
    left_join(., 
              wkr.df %>% group_by(TubeNo) %>% 
    summarise(el_m=mean(el_m), 
              across(matches(Y_var), 
                     .fns=list(mn=mean, sd=sd),
                     .names="obs_{.fn}")))
  
  p <- ggplot(post, aes(x=el_m, y=mu)) + 
    stat_interval(.width=c(0.5, 0.8, 0.9, 0.95, 0.99), 
                  point_interval=median_hdi, size=1) +
    geom_point(data=post %>% group_by(TubeNo) %>% slice_head(n=1), 
               aes(y=obs_mn), shape=1, size=1, alpha=0.5) +
    stat_smooth(method="lm", formula=y~x, size=0.25) + 
    facet_wrap(~SPECIESID, scales="free") + 
    scale_colour_brewer("HPDI", type="seq") +
    labs(x="Elevation (m)", y="Colony mean", title=Y_var) +
    theme(legend.position=c(0.9, 0.08), 
          legend.key.height=unit(0.1, "cm"))
  ggsave(paste0(fig_dir, "el_mu_", Y_var, std_ext, ".png"), p, 
         width=9, height=9)
  
  p <- ggplot(post, aes(x=el_m, y=sigma)) + 
    stat_interval(.width=c(0.5, 0.8, 0.9, 0.95, 0.99), 
                  point_interval=median_hdi, size=1) +
    geom_point(data=post %>% group_by(TubeNo) %>% slice_head(n=1), 
               aes(y=obs_sd), shape=1, size=1, alpha=0.5) +
    stat_smooth(method="lm", formula=y~x, size=0.25) + 
    facet_wrap(~SPECIESID, scales="free") + 
    scale_colour_brewer("HPDI", type="seq") +
    scale_y_log10() +
    labs(x="Elevation (m)", y="Colony log(sd)", title=Y_var) +
    theme(legend.position=c(0.9, 0.08), 
          legend.key.height=unit(0.1, "cm"))
  ggsave(paste0(fig_dir, "el_sd_", Y_var, std_ext, ".png"), p, 
         width=9, height=9)
  
  p <- post %>% group_by(TubeNo) %>% sample_n(1e3) %>%
    mutate(response=map2(mu, sigma, ~rnorm(10, .x, .y))) %>%
    unnest(cols=response) %>%
    ggplot(aes(x=el_m, y=response)) + 
    stat_interval(.width=c(0.5, 0.6, 0.7, 0.8, 0.9, 0.95), 
                  point_interval=median_hdi, size=0.8) +
    stat_smooth(method="lm", formula=y~x, size=0.4, 
                linetype=2, colour="grey30") + 
    scale_colour_viridis_d("HPDI", option="E") +
    facet_wrap(~SPECIESID, scales="free") + 
    labs(x="Elevation (m)", y=Y_var, 
         title=paste("Posterior colony-level distributions:", Y_var)) +
    theme(legend.position=c(0.9, 0.08), 
          legend.key.height=unit(0.1, "cm"))
  ggsave(paste0(fig_dir, "el_y_", Y_var, std_ext, ".png"), p, 
         width=9, height=9)
}
```




----------




# Elevation  
First, we can establish whether there are any detectable changes across elevations in each trait, either in the mean or the standard deviation within each colony. 

```{r el_b_eff}
library(tidyverse); library(tidybayes)
theme_set(theme_bw() + theme(panel.grid.minor=element_blank(),
                             panel.grid.major=element_line(colour="gray90",
                                                           size=0.25)))

b_el.df <- dir("eda/el", "csv", full.names=T) %>% map_dfr(read_csv)

focus_traits <- c("Scape Proportion",
                  "Interocular Distance", 
                  "Leg Length (hind)",
                  "Pronotal Expansion",
                  "Head Shape",
                  "Head Width", 
                  "Head Length",
                  "Webers Length", 
                  "Color (lightness)")

filter(b_el.df, covariate != "Intercept") %>% 
  mutate(covariate=cov_names$full[match(covariate, cov_names$orig)],
         dpar=if_else(dpar=="mu", "mean_colony", "sd_colony"),
         Trait=trait_match$full[match(Yvar, trait_match$orig)],
         Trait=factor(Trait, levels=focus_traits),
         TraitType=trait_match$relative[match(Yvar, trait_match$orig)]) %>%
  ggplot(aes(x=.value, xmin=.lower, xmax=.upper, y=Trait, colour=TraitType)) + 
  geom_vline(xintercept=0, colour="grey") +
  geom_pointinterval(position=position_dodge(width=0.5), fatten_point=0.8) +
  scale_colour_brewer(type="qual", palette=2, labels=c("Absolute", "Relative")) +
  facet_grid(dpar~covariate) + 
  labs(x="Posterior distribution", y="", title="Elevation")
ggsave("eda/el/ALL_b_eff.png", width=7, height=5)
```



----------





# Specific hypotheses  

```{r hyp_b_eff}
library(tidyverse); library(tidybayes)
theme_set(theme_bw() + theme(panel.grid.minor=element_blank(),
                             panel.grid.major=element_line(colour="gray90",
                                                           size=0.25)))

b_hyp.df <- dir("eda/hyp", "csv", full.names=T) %>% map_dfr(read_csv)

focus_traits <- c("Scape Proportion",
                  "Interocular Distance", 
                  "Leg Length (hind)",
                  "Pronotal Expansion",
                  "Head Shape",
                  "Head Width", 
                  "Head Length",
                  "Webers Length", 
                  "Color (lightness)")

filter(b_hyp.df, covariate != "Intercept") %>% 
  mutate(covariate=cov_names$full[match(covariate, cov_names$orig)],
         dpar=if_else(dpar=="mu", "mean_colony", "sd_colony"),
         Trait=trait_match$full[match(Yvar, trait_match$orig)],
         Trait=factor(Trait, levels=focus_traits),
         TraitType=trait_match$relative[match(Yvar, trait_match$orig)]) %>%
  ggplot(aes(x=.value, xmin=.lower, xmax=.upper, y=Trait, colour=TraitType)) + 
  geom_vline(xintercept=0, colour="grey") +
  geom_pointinterval(position=position_dodge(width=0.5), fatten_point=0.8) +
  scale_colour_brewer(type="qual", palette=2, labels=c("Absolute", "Relative")) +
  facet_grid(dpar~covariate) + 
  theme(legend.position="bottom") +
  labs(x="Posterior distribution", y="", title="Theoretical or empirical justification")
ggsave("eda/hyp/ALL_b_eff.png", width=8, height=4.5)
```





----------





# Climatic variables 

```{r clim_b_eff}
library(tidyverse); library(tidybayes)
theme_set(theme_bw() + theme(panel.grid.minor=element_blank(),
                             panel.grid.major=element_line(colour="gray90",
                                                           size=0.25)))

b_clim.df <- dir("eda/climate", "csv", full.names=T) %>% map_dfr(read_csv)

focus_traits <- c("Scape Proportion",
                  "Interocular Distance", 
                  "Leg Length (hind)",
                  "Pronotal Expansion",
                  "Head Shape",
                  "Head Width", 
                  "Head Length",
                  "Webers Length", 
                  "Color (lightness)")

filter(b_clim.df, covariate != "Intercept") %>% 
  mutate(covariate=cov_names$full[match(covariate, cov_names$orig)],
         dpar=if_else(dpar=="mu", "mean_colony", "sd_colony"),
         Trait=trait_match$full[match(Yvar, trait_match$orig)],
         Trait=factor(Trait, levels=focus_traits),
         TraitType=trait_match$relative[match(Yvar, trait_match$orig)]) %>%
  ggplot(aes(x=.value, xmin=.lower, xmax=.upper, y=Trait, colour=TraitType)) + 
  geom_vline(xintercept=0, colour="grey") +
  geom_pointinterval(position=position_dodge(width=0.5), fatten_point=0.8) +
  scale_colour_brewer(type="qual", palette=2, labels=c("Absolute", "Relative")) +
  facet_grid(dpar~covariate) + 
  labs(x="Posterior distribution", y="", title="Climatic variables")
ggsave("eda/climate/ALL_b_eff.png", width=9, height=5)
```





----------





# All variables 

```{r clim_b_eff}
library(tidyverse); library(tidybayes)
theme_set(theme_bw() + theme(panel.grid.minor=element_blank(),
                             panel.grid.major=element_line(colour="gray90",
                                                           size=0.25)))

b_all.df <- dir("eda/all", "csv", full.names=T) %>% map_dfr(read_csv)

focus_traits <- c("Scape Proportion",
                  "Interocular Distance", 
                  "Leg Length (hind)",
                  "Pronotal Expansion",
                  "Head Shape",
                  "Head Width", 
                  "Head Length",
                  "Webers Length", 
                  "Color (lightness)")

filter(b_all.df, covariate != "Intercept") %>% 
  mutate(covariate=cov_names$full[match(covariate, cov_names$orig)],
         dpar=if_else(dpar=="mu", "mean_colony", "sd_colony"),
         Trait=trait_match$full[match(Yvar, trait_match$orig)],
         Trait=factor(Trait, levels=focus_traits),
         TraitType=trait_match$relative[match(Yvar, trait_match$orig)]) %>%
  ggplot(aes(x=.value, xmin=.lower, xmax=.upper, y=Trait, colour=TraitType)) + 
  geom_vline(xintercept=0, colour="grey") +
  geom_pointinterval(position=position_dodge(width=0.5), fatten_point=0.8) +
  scale_colour_brewer(type="qual", palette=2, labels=c("Absolute", "Relative")) +
  facet_grid(dpar~covariate) + 
  labs(x="Posterior distribution", y="", title="All variables")
ggsave("eda/all/ALL_b_eff.png", width=10, height=5)
```
